---
title: "Team Money Dataset"
subtitle: "Pacific Federal Loan Campaign"
author: "Charlie, Arya, Steven, Ryan (Team Money)"
format:
  html:
    df-print: paged
    number-sections: true
    toc: true
    toc-depth: 2
    code-tools: true
    code-overflow: wrap
    embed-resources: true
execute:
  echo: true
  warning: false
  message: false
  cache: false
  error: false
---

```{r}
# Load libs
library(tidyverse)

# set seed
set.seed(310)

# Printing / numeric display
options(width = 70)
options(scipen = 99)

# Chunk defaults (optional, but keeps figures consistent)
knitr::opts_chunk$set(
  fig.width  = 5,
  fig.height = 4,
  fig.align  = "center"
  )
```

# Load the Data

```{r}
loans <- read_csv("data/pacific_federal_loan_campaign.csv")
```

# Overview

```{r}
# Dimensions: rows x columns
dim(loans)
```

```{r}
glimpse(loans)
```

```{r}
head(loans)
```

# Summary Statistics

```{r}
summary(loans)
```

# Data Cleaning

```{r}
loans_clean <- 
  loans %>% 
  mutate(
    Education = factor(Education),
    Personal_Loan = factor(Personal_Loan),
    Securities_Account = factor(Securities_Account),
    CD_Account = factor(CD_Account),
    Online = factor(Online),
    CreditCard = factor(CreditCard)
  )
```

# Model

```{r}
library(caret)
library(rsample)
set.seed(310)
loans_split <- initial_split(loans_clean, prop=0.8)
loans_train <- training(loans_split)
loans_test  <- testing(loans_split)
logit1 <- glm(Personal_Loan ~ Age + Experience + Income + Education + Mortgage, 
              data=loans_train, 
              family = binomial)

summary(logit1)

scores_train <- predict(logit1, newdata=loans_train, type="response")
results_train <- 
  loans_train %>% 
    mutate(scores = scores_train)


```

# Visualizations

```{r}
library(ggplot2)

# Visualization 1: This visualization examines the relationship between annual income and the propensity to accept a personal loan, segmented by education level.

ggplot(loans_train, aes(x = Income, fill = as.factor(Personal_Loan))) +
  geom_density(alpha = 0.5) +
  facet_wrap(~Education) +
  labs(title = "Income Distribution by Loan Acceptance",
       subtitle = "Faceted by Education Level",
       x = "Annual Income (Thousands $)",
       fill = "Personal Loan") +
  theme_minimal()


# Visualization 2: This plot visualizes the predicted probabilities against income, which we believe could be a very significant predictor. 
ggplot(results_train, aes(x = Income, y = scores, color = Personal_Loan)) +
  geom_point(alpha = 0.6) +
  geom_hline(yintercept = 0.5, color="red", linetype="dashed") + 
  labs(title = "Predicted Loan Probability vs. Income",
       x = "Income",
       y = "Predicted Probability of Accepting Loan") +
  theme_minimal()
```

# Plan

1.  More in depth data cleaning (looking at data for outliers and removing them along with missing values)
2.  Adjusting model to find predictors that have the largest prediction impact (rather than just using variables we think may be useful)
3.  Experiment with numerous types of models to find which one best pertains to our question and dataset
4.  Predicting the probabilities of a certain person accepting the loan offer or not
5.  Deciding what metrics are most useful to measure how impactful our model is
6.  Creating visualizations that paint a picture to stakeholders of the type of clients they should be targeting with the new campaign
7.  Refining our model(s)

# Roadblocks/Questions

Deciding the type of model we want to use has been a roadblock. We will continue using what we learned in class (such as random forests) to experiment. Our preliminary model is a logistic regression.
