---
title: "Team Money Dataset"
subtitle: "Pacific Federal Loan Campaign"
author: "Charlie, Arya, Steven, Ryan (Team Money)"
format:
  html:
    df-print: paged
    number-sections: true
    toc: true
    toc-depth: 2
    code-tools: true
    code-overflow: wrap
    embed-resources: true
execute:
  echo: true
  warning: false
  message: false
  cache: false
  error: false
---

```{r}
# Load libs
library(tidyverse)

# set seed
set.seed(310)

# Printing / numeric display
options(width = 70)
options(scipen = 99)

# Chunk defaults (optional, but keeps figures consistent)
knitr::opts_chunk$set(
  fig.width  = 5,
  fig.height = 4,
  fig.align  = "center"
  )
```

# Business Problem

Pacific Federal is a regional bank whose liability (deposit) customers outnumber its asset (loan) customers. The retail banking division recently ran a campaign offering personal loans to its deposit holders, but fewer than 10% accepted. The Chief Retail Officer wants to understand what distinguishes customers who accepted the loan from those who didn't, and build a system to identify high-potential targets for the next campaign round.

# Load the Data

```{r}
loans <- read_csv("data/pacific_federal_loan_campaign.csv")
```

# Overview

```{r}
# Dimensions: rows x columns
dim(loans)
```

```{r}
glimpse(loans)
```

```{r}
head(loans)
```

# Summary Statistics

```{r}
summary(loans)
```

-   9.6% of users accepted the loan offer

-   Average household size is around 2.396

-   10.4% of users have a securities account

-   6% of users have a CD

-   29.4% of users have an active credit card

# Data Cleaning

```{r}
loans_clean <- 
  loans %>% 
  filter(Experience >= 0) %>%
  mutate(
    Education = factor(Education, 
                       levels = c(1, 2, 3),
                       labels = c("Undergraduate", "Graduate", "Professional")),
    Personal_Loan = factor(Personal_Loan),
    Securities_Account = factor(Securities_Account),
    CD_Account = factor(CD_Account),
    Online = factor(Online),
    CreditCard = factor(CreditCard),
    Income_Per_Family_Member = (Income / Family), 
    Monthly_Income = Income * 12,
    Monthly_CC_Spending_to_Income = CCAvg / Monthly_Income,
    income_over_100k = ifelse(Income >= 100, 1, 0)
  )
```

-   Entries with negative years of experience removed

-   Education changed to factor with labels in order to make visualizations with this predictor easier to interpret

-   Income_Per_Family_Member created to find contribution of each member (gives insight into how much is going to expenses if family is larger)

-   Monthly_CC_Spending_to_Income created to see how much of monthly income goes towards credit card

# Data Splitting

```{r}
library(caret)
library(rsample)

set.seed(310)
loans_split <- initial_split(loans_clean, prop=0.8)
loans_train <- training(loans_split)
loans_test  <- testing(loans_split)
```

# Model 1

```{r}
library(caret)
library(rsample)

set.seed(310)
loans_split <- initial_split(loans_clean, prop=0.8)
loans_train <- training(loans_split)
loans_test  <- testing(loans_split)
logit1 <- glm(Personal_Loan ~ Experience + Income + Education + Family + CD_Account,
              data=loans_train, 
              family = binomial)

summary(logit1)

```

-   Mortgage removed due to low significance

-   Age removed due to likely high correlation with experience

# Predict Model 1

```{r}

scores_train <- predict(logit1, newdata=loans_train, type="response")
results_train <- 
  loans_train %>% 
    mutate(scores_tr = scores_train)

scores_test <- predict(logit1, newdata=loans_test, type="response")
results_test <- 
  loans_test %>% 
    mutate(scores_te = scores_test)

```

# Report Results Model 1

```{r}
pred_class <- ifelse(scores_test > 0.63, 1, 0)


actual <- factor(loans_test$Personal_Loan,
          levels = c(0,1))

predicted <- factor(pred_class,
              levels = c(0,1))

cm <- confusionMatrix(predicted,
      actual,
      positive = "1")

cm
```

-   Initially started with a 90% cutoff because acceptance rate is around 10%

    -   Lowered gradually until 0.62

# ROC/AUC Model 1 

```{r}
library(pROC)
roc <- roc(actual, scores_test)

# AUC

auc(roc)

# Plot ROC Curve

plot(roc)
```

# Visualizations

```{r}
library(ggplot2)

# Visualization 1: Probability of accepting a personal loan by education level

ggplot(loans_train,
       aes(x = Education,
           fill = as.factor(Personal_Loan))) +
  geom_bar(position = "fill") +
  scale_y_continuous(labels = scales::percent) +
  labs(
    title = "Probability of Accepting a Personal Loan by Education Level",
    x = "Education Level",
    y = "Probability",
    fill = "Personal Loan"
  ) +
  theme_minimal()


# Visualization 2: This plot visualizes the predicted probabilities against income, which we believe could be a very significant predictor. 
ggplot(results_train, aes(x = Income, y = scores_train, color = Personal_Loan)) +
  geom_point(alpha = 0.6) +
  geom_hline(yintercept = 0.5, color="red", linetype="dashed") + 
  labs(title = "Predicted Loan Probability vs. Income",
       x = "Income",
       y = "Predicted Probability of Accepting Loan") +
  theme_minimal()

# Visualization 3: This plots visualizes the amount of users that took the personal loan based on if they had a securities account, cd, or credit card.

library(dplyr)

plot_data <- data.frame(
  Type = c("Securities Account", "CD Account", "Credit Card", "All Three"),
  Percent = c(
    sum(loans_clean$Personal_Loan == 1 & loans_clean$Securities_Account == 1) /
      sum(loans_clean$Securities_Account == 1),
    sum(loans_clean$Personal_Loan == 1 & loans_clean$CD_Account == 1) /
      sum(loans_clean$CD_Account == 1),
    sum(loans_clean$Personal_Loan == 1 & loans_clean$CreditCard == 1) /
      sum(loans_clean$CreditCard == 1),
    sum(loans_clean$Personal_Loan == 1 &
          loans_clean$Securities_Account == 1 &
          loans_clean$CD_Account == 1 &
          loans_clean$CreditCard == 1) /
      sum(loans_clean$Securities_Account == 1 &
            loans_clean$CD_Account == 1 &
            loans_clean$CreditCard == 1)
  )
)
ggplot(plot_data, aes(Type, Percent, fill = Type)) +
  geom_col() +
  geom_text(aes(label = scales::percent(Percent)), vjust = -0.3) +
  scale_y_continuous(labels = scales::percent) +
  labs(
    title = "Percent of Users Taking Personal Loans",
    x = "",
    y = "Percent Taking Loan"
  ) +
  theme_minimal()

```

# Model 2

```{r}
library(glmnet)
library(caret)
library(dplyr)

x_train <- model.matrix(
  Personal_Loan ~ Experience + Income + Education +
  Monthly_CC_Spending_to_Income + CD_Account + Income_Per_Family_Member
  + CreditCard + income_over_100k,
  data = loans_train )[, -1]

y_train <- loans_train$Personal_Loan

x_test <- model.matrix(
  Personal_Loan ~ Experience + Income + Education +
  Monthly_CC_Spending_to_Income + CD_Account + Income_Per_Family_Member
  + CreditCard + income_over_100k,
  data = loans_test
)[, -1]

lasso_model <- cv.glmnet(
  x_train,
  y_train,
  family = "binomial",
  alpha = 1
)

coef(lasso_model, s = "lambda.min")


```

# Predict Model 2

```{r}
lasso_probs <- predict(
  lasso_model,
  s = "lambda.min",
  newx = x_test,
  type = "response"
)
```

# Report Results Model 2

```{r}

lasso_pred <- ifelse(lasso_probs > 0.6, 1, 0)

actual_lasso <- factor(loans_test$Personal_Loan,
                 levels = c(0,1))

predicted_lasso <- factor(lasso_pred,
                    levels = c(0,1))

cm_lasso <- confusionMatrix(predicted_lasso,
                      actual_lasso,
                      positive = "1")

cm_lasso

```

# ROC/AUC Model 2

```{r}

library(pROC)

roc_lasso <- roc(actual_lasso, as.numeric(lasso_probs))
auc(roc_lasso)
plot(roc_lasso)

```

# Interpretation

Across both models, the predictors that mattered most were Income, Education level, and CD account ownership. In Model 1, Income was very significant (p < 2e-16) with a coefficient of 0.063, so every extra thousand dollars of income raises the chance a customer accepts the loan. Education had an even bigger effect: Graduate and Professional customers had log-odds about 3.6 higher than Undergraduate customers, holding everything else equal. CD_Account came in at 2.76, meaning customers who already have a CD are much more likely to take the loan too.

These results make sense for the business. Higher-income customers have room in their budget to take on a loan, and the bank can lend to them with less risk. More educated customers usually earn more over time and may be more used to working with debt products. CD holders are already invested with the bank, so they would be more open to a new product from the same place.

Model 2 told a similar story but added two new pieces. The income_over_100k feature we built came in at 2.46, which says the relationship between income and acceptance is not perfectly linear and something extra kicks in once a customer crosses the 100k mark. CreditCard came in at -0.95, meaning customers who already have a credit card are a little less likely to accept the loan. That was a small surprise. Our best guess is they already have access to short-term credit, so a separate loan is less appealing.

Experience came back as not significant in Model 1 (p ≈ 0.75). We had kept it in to capture career stage, but it did not add much once Income was in the model. Family was positive in Model 1 while Income_Per_Family_Member was negative in Model 2, which looked like a contradiction at first. The signs actually agree: for a given income, a bigger family means each person has less to work with, and those households are more open to a loan. The lasso also did not drop any predictors at the lambda that minimized CV error, so it acted like a regularized logit rather than doing real variable selection.

The bigger issue is sensitivity. Model 1 looks great at 95% accuracy, but at the 0.63 cutoff it only catches 53% of the customers who actually took the loan. Model 2 brings that to 71%, which is a real jump, but we are still missing close to 30% of the people the bank wants to find. With only 9.6% accepting, accuracy alone is misleading. A model that predicts "no" for everyone would land around 90% accuracy and be useless.

A few limitations worth flagging. We only have one campaign of data, so we cannot say whether these patterns hold up over time or after the bank changes its product mix. We dropped ZIPCode, Online, and Securities_Account before modeling, and there might be signal in those. We also did not try interaction terms like Income times Education, which might have caught the nonlinearity above 100k more cleanly.

# Conclusion & Recommendation

We recommend Model 2, the lasso. It beat Model 1 on AUC (0.970 vs 0.962) and was clearly better on sensitivity at the cutoff we chose (71% vs 53%), which is the number we care about most here. The bank wants to find customers who will say yes more than it wants to avoid reaching out to people who say no. Outreach is cheap, and a missed loan costs real money in lost interest. Because of that, we would push the cutoff a bit below 0.6 in practice. That raises sensitivity a little more at the cost of a few extra false positives, and the trade favors the bank.

For the Chief Retail Officer, the takeaways come straight out of those coefficients. The first priority should be high-income deposit customers, especially anyone above 100k, since the income_over_100k indicator alone shifted the odds by a lot. Graduate and Professional degree customers convert at much higher rates than Undergraduate customers, so the campaign could lean into that audience in the messaging or product framing. CD account holders are a strong second group: they are already invested with the bank and the model treats them as warm leads. Family size also matters once income is controlled for, so households with several dependents and moderate incomes look more loan-receptive than the raw averages suggest. Customers who already carry a credit card look like cooler leads than they otherwise appear, so the outreach budget could go elsewhere.

If we had more time and data, there are a few things we would want to do. The first is to get an actual cost estimate from the marketing team for outreach versus a missed loan, then re-tune the cutoff to minimize expected dollar loss instead of picking a threshold by eye. The second is to try a tree-based model like random forest or XGBoost, since logit and lasso both assume linearity in the log-odds and acceptance probably has thresholds and interactions those models cannot pick up. We would also want to pull data from earlier campaigns to check whether the patterns hold up over time. Finally, if the bank has behavioral data like online banking activity or app logins, bringing that in would probably help with customers who do not have obvious income or CD signals.

