---
title: "Team Money Dataset"
subtitle: "Pacific Federal Loan Campaign"
author: "Charlie, Arya, Steven, Ryan (Team Money)"
format:
  html:
    df-print: paged
    number-sections: true
    toc: true
    toc-depth: 2
    code-tools: true
    code-overflow: wrap
    embed-resources: true
execute:
  echo: false
  warning: false
  message: false
  cache: false
  error: false
---

```{r}
# Load libs
library(tidyverse)

# set seed
set.seed(310)

# Printing / numeric display
options(width = 70)
options(scipen = 99)

# Chunk defaults (optional, but keeps figures consistent)
knitr::opts_chunk$set(
  fig.width  = 5,
  fig.height = 4,
  fig.align  = "center"
  )
```

# Introduction & Business Problem

Pacific Federal is a regional bank with more deposit customers than loan customers, and the retail banking division wants to convert more of those deposit holders into personal loan customers. The most recent campaign offered a personal loan to existing deposit holders and fewer than 10% accepted. That low conversion rate means the bank is spending outreach effort on a lot of customers who never planned to take the loan, while potentially missing better leads inside the same base.

The decision-maker is the Chief Retail Officer, who needs to decide who to target in the next campaign round. A predictive model helps directly here. If we can estimate each customer's probability of accepting a loan based on what the bank already knows about them, the CRO can rank the deposit base from most to least likely to accept and focus the budget on the top of that list. Even a moderate lift over the 9.6% baseline acceptance rate would translate into noticeably more loans booked at a similar campaign cost.

# Data Summary & Cleaning

```{r}
loans <- read_csv("data/pacific_federal_loan_campaign.csv")
```

## Summary Statistics

The dataset has 5,000 customer records with 14 columns covering demographics (Age, Experience, Education, Family, ZIP), financial details (Income, CCAvg, Mortgage), and existing product ownership (Securities Account, CD Account, Online banking, Credit Card). The outcome variable is Personal_Loan, a binary flag for whether the customer accepted the loan offer in the previous campaign.

The most important thing to flag up front is the class imbalance: only 9.6% of customers accepted, so any model we build needs to be evaluated with that imbalance in mind. The customer base skews older and more established, with a mean age of 45 and 20 years of experience on average. Income ranges from 8k to 224k with a mean around 74k, which gives the models room to separate high- and low-likelihood customers. Existing product ownership at the bank is generally low: around 10% of customers hold a securities account, 6% hold a CD, and 29% carry a bank credit card. We also noticed that 52 records have negative years of experience, which is impossible and gets cleaned in the next section.

Two patterns in the raw data jumped out during exploration and are worth flagging before the cleaning and modeling steps.

```{r}
library(ggplot2)

ggplot(loans %>% filter(Experience >= 0),
       aes(x = factor(Education, levels = c(1, 2, 3),
                      labels = c("Undergraduate", "Graduate", "Professional")),
           fill = as.factor(Personal_Loan))) +
  geom_bar(position = "fill") +
  scale_y_continuous(labels = scales::percent) +
  labs(
    title = "Probability of Accepting a Personal Loan by Education Level",
    x = "Education Level",
    y = "Probability",
    fill = "Personal Loan"
  ) +
  theme_minimal()
```

Graduate and Professional customers accept at roughly 14% versus 5% for Undergraduate, a sharp gap that flagged Education as a likely strong predictor.

```{r}
library(dplyr)

plot_data <- data.frame(
  Type = c("Securities Account", "CD Account", "Credit Card", "All Three"),
  Percent = c(
    sum(loans$Personal_Loan == 1 & loans$Securities_Account == 1) /
      sum(loans$Securities_Account == 1),
    sum(loans$Personal_Loan == 1 & loans$CD_Account == 1) /
      sum(loans$CD_Account == 1),
    sum(loans$Personal_Loan == 1 & loans$CreditCard == 1) /
      sum(loans$CreditCard == 1),
    sum(loans$Personal_Loan == 1 &
          loans$Securities_Account == 1 &
          loans$CD_Account == 1 &
          loans$CreditCard == 1) /
      sum(loans$Securities_Account == 1 &
            loans$CD_Account == 1 &
            loans$CreditCard == 1)
  )
)
ggplot(plot_data, aes(Type, Percent, fill = Type)) +
  geom_col() +
  geom_text(aes(label = scales::percent(Percent)), vjust = -0.3) +
  scale_y_continuous(labels = scales::percent) +
  labs(
    title = "Percent of Users Taking Personal Loans",
    x = "",
    y = "Percent Taking Loan"
  ) +
  theme_minimal()
```

CD account holders accept at about 46%, nearly five times the 9.6% baseline, while credit card and securities holders sit close to baseline.

## Data Cleaning

```{r}
loans_clean <-
  loans %>%
  filter(Experience >= 0) %>%
  mutate(
    Education = factor(Education,
                       levels = c(1, 2, 3),
                       labels = c("Undergraduate", "Graduate", "Professional")),
    Personal_Loan = factor(Personal_Loan),
    Securities_Account = factor(Securities_Account),
    CD_Account = factor(CD_Account),
    Online = factor(Online),
    CreditCard = factor(CreditCard)
  )
```

-   Entries with negative years of experience removed

-   Education changed to factor with labels in order to make visualizations with this predictor easier to interpret

# Feature Engineering

Beyond the basic cleaning, we created four engineered features to capture relationships we thought the raw columns would not pick up on their own.

```{r}
loans_clean <-
  loans_clean %>%
  mutate(
    Income_Per_Family_Member = Income / Family,
    Monthly_Income = Income * 12,
    Monthly_CC_Spending_to_Income = CCAvg / Monthly_Income,
    income_over_100k = ifelse(Income >= 100, 1, 0)
  )
```

Income_Per_Family_Member came from the thought that household income tells us what the family earns but not how stretched that money is, so we divided income by family size to get a per-person figure. We expected a positive coefficient, but the lasso gave it a small negative one. That actually makes sense after the fact: for a fixed income, larger families have more financial need and may be more open to a loan.

Monthly_Income is just Income multiplied by 12. Most household financial decisions get framed in monthly terms, so we built a monthly version mainly to feed into the next feature.

Monthly_CC_Spending_to_Income puts credit card spending in context. Someone charging 2k a month on 200k of income is in a very different spot than someone charging the same on 40k, so dividing CCAvg by Monthly_Income gives a ratio that should pick up customers running close to their limits, who might be more receptive to a personal loan to consolidate.

income_over_100k is a binary indicator for customers earning at least 100k. During exploration we noticed acceptance rates climbed sharply in the upper income tiers rather than gradually, so a binary lets the model pick up that jump without assuming the effect is linear in income. The lasso gave it a coefficient of 2.46, confirming a real threshold effect Income alone was missing.

A few things did not pay off. Monthly_Income on its own added nothing once Income was in the model, since a linear model treats scalar transformations as redundant. We also considered Mortgage in Model 1 but dropped it because it came out insignificant, and we dropped Age because it overlapped heavily with Experience.

# Modeling & Results

## Data Splitting

```{r}
library(caret)
library(rsample)

set.seed(310)
loans_split <- initial_split(loans_clean, prop=0.8)
loans_train <- training(loans_split)
loans_test  <- testing(loans_split)
```

## Model 1

```{r}
library(caret)
library(rsample)

set.seed(310)
loans_split <- initial_split(loans_clean, prop=0.8)
loans_train <- training(loans_split)
loans_test  <- testing(loans_split)
logit1 <- glm(Personal_Loan ~ Experience + Income + Education + Family + CD_Account,
              data=loans_train, 
              family = binomial)

summary(logit1)

```

-   Mortgage removed due to low significance

-   Age removed due to likely high correlation with experience

```{r}

scores_train <- predict(logit1, newdata=loans_train, type="response")
results_train <- 
  loans_train %>% 
    mutate(scores_tr = scores_train)

scores_test <- predict(logit1, newdata=loans_test, type="response")
results_test <- 
  loans_test %>% 
    mutate(scores_te = scores_test)

```

```{r}
pred_class <- ifelse(scores_test > 0.63, 1, 0)


actual <- factor(loans_test$Personal_Loan,
          levels = c(0,1))

predicted <- factor(pred_class,
              levels = c(0,1))

cm <- confusionMatrix(predicted,
      actual,
      positive = "1")

cm
```

-   Initially started with a 90% cutoff because acceptance rate is around 10%

    -   Lowered gradually until 0.62

```{r}
library(pROC)
roc <- roc(actual, scores_test)

# AUC

auc(roc)

# Plot ROC Curve

plot(roc)
```

```{r}
ggplot(results_train, aes(x = Income, y = scores_train, color = Personal_Loan)) +
  geom_point(alpha = 0.6) +
  geom_hline(yintercept = 0.5, color = "red", linetype = "dashed") +
  labs(title = "Predicted Loan Probability vs. Income",
       x = "Income",
       y = "Predicted Probability of Accepting Loan") +
  theme_minimal()
```

Predicted acceptance stays near zero until income hits about 100k and then rises sharply, which is what motivated the income_over_100k feature in Model 2.

## Model 2

```{r}
library(glmnet)
library(caret)
library(dplyr)

x_train <- model.matrix(
  Personal_Loan ~ Experience + Income + Education +
  Monthly_CC_Spending_to_Income + CD_Account + Income_Per_Family_Member
  + CreditCard + income_over_100k,
  data = loans_train )[, -1]

y_train <- loans_train$Personal_Loan

x_test <- model.matrix(
  Personal_Loan ~ Experience + Income + Education +
  Monthly_CC_Spending_to_Income + CD_Account + Income_Per_Family_Member
  + CreditCard + income_over_100k,
  data = loans_test
)[, -1]

lasso_model <- cv.glmnet(
  x_train,
  y_train,
  family = "binomial",
  alpha = 1
)

coef(lasso_model, s = "lambda.min")


```

```{r}
lasso_probs <- predict(
  lasso_model,
  s = "lambda.min",
  newx = x_test,
  type = "response"
)
```

```{r}

lasso_pred <- ifelse(lasso_probs > 0.6, 1, 0)

actual_lasso <- factor(loans_test$Personal_Loan,
                 levels = c(0,1))

predicted_lasso <- factor(lasso_pred,
                    levels = c(0,1))

cm_lasso <- confusionMatrix(predicted_lasso,
                      actual_lasso,
                      positive = "1")

cm_lasso

```

```{r}

library(pROC)

roc_lasso <- roc(actual_lasso, as.numeric(lasso_probs))
auc(roc_lasso)
plot(roc_lasso)

```

## Model Evaluation & Comparison

We compared the two models on the metrics that matter for a classification problem with this kind of class imbalance:

| Metric | Model 1 (Logit) | Model 2 (Lasso) |
|---|---|---|
| Cutoff | 0.63 | 0.6 |
| Accuracy | 95.2% | 96.5% |
| Sensitivity | 52.8% | 70.8% |
| Specificity | 99.3% | 99.0% |
| AUC | 0.962 | 0.970 |

Model 2 beats Model 1 on AUC and meaningfully on sensitivity, while giving up almost nothing on specificity. The accuracy numbers look similar, but accuracy is misleading here. With only 9.6% of customers accepting the loan, a model that predicts "no" for everyone would land around 90% accuracy and find zero takers. Sensitivity is the metric that matters because it measures how well the model identifies the customers the bank actually wants to find.

The cost of each error type also matters for picking a cutoff. A false positive means the bank reaches out to a customer who will not accept, and the cost is the marketing spend on that one contact. A false negative means the bank misses a customer who would have accepted, and the cost is the foregone interest income on a loan that does not get made. The interest on a personal loan over its lifetime is far larger than the cost of a single outreach contact, so the false negative cost dominates. That asymmetry is the reason we picked cutoffs on the lower end of what still kept specificity high, and it is the main reason we recommend Model 2 over Model 1.

# Interpretation

Across both models, the predictors that mattered most were Income, Education level, and CD account ownership. In Model 1, Income was very significant (p < 2e-16) with a coefficient of 0.063, so every extra thousand dollars of income raises the chance a customer accepts the loan. Education had an even bigger effect: Graduate and Professional customers had log-odds about 3.6 higher than Undergraduate customers, holding everything else equal. CD_Account came in at 2.76, meaning customers who already have a CD are much more likely to take the loan too.

These results make sense for the business. Higher-income customers have room in their budget to take on a loan, and the bank can lend to them with less risk. More educated customers usually earn more over time and may be more used to working with debt products. CD holders are already invested with the bank, so they would be more open to a new product from the same place.

Model 2 told a similar story but added two new pieces. The income_over_100k feature we built came in at 2.46, which says the relationship between income and acceptance is not perfectly linear and something extra kicks in once a customer crosses the 100k mark. CreditCard came in at -0.95, meaning customers who already have a credit card are a little less likely to accept the loan. That was a small surprise. Our best guess is they already have access to short-term credit, so a separate loan is less appealing.

Experience came back as not significant in Model 1 (p ≈ 0.75). We had kept it in to capture career stage, but it did not add much once Income was in the model. Family was positive in Model 1 while Income_Per_Family_Member was negative in Model 2, which looked like a contradiction at first. The signs actually agree: for a given income, a bigger family means each person has less to work with, and those households are more open to a loan. The lasso also did not drop any predictors at the lambda that minimized CV error, so it acted like a regularized logit rather than doing real variable selection.

Even with the better sensitivity we get from Model 2, the model still misses close to 30% of the customers who actually took the loan. That is the most honest limitation of the work and a reminder that this problem is hard to solve with the variables we had.

A few limitations worth flagging. We only have one campaign of data, so we cannot say whether these patterns hold up over time or after the bank changes its product mix. We dropped ZIPCode, Online, and Securities_Account before modeling, and there might be signal in those. We also did not try interaction terms like Income times Education, which might have caught the nonlinearity above 100k more cleanly.

# Conclusion & Recommendation

We recommend Model 2, the lasso. As laid out in the evaluation section, it cleanly outperforms Model 1 on the metrics that matter most for this problem without giving up meaningful ground on specificity. Given the cost asymmetry between false positives and false negatives, we would actually push the cutoff slightly below 0.6 in production to capture a few more positives, since the extra outreach is cheaper to absorb than additional missed customers.

For the Chief Retail Officer, the takeaways come straight out of those coefficients. The first priority should be high-income deposit customers, especially anyone above 100k, since the income_over_100k indicator alone shifted the odds by a lot. Graduate and Professional degree customers convert at much higher rates than Undergraduate customers, so the campaign could lean into that audience in the messaging or product framing. CD account holders are a strong second group: they are already invested with the bank and the model treats them as warm leads. Family size also matters once income is controlled for, so households with several dependents and moderate incomes look more loan-receptive than the raw averages suggest. Customers who already carry a credit card look like cooler leads than they otherwise appear, so the outreach budget could go elsewhere.

If we had more time and data, there are a few things we would want to do. The first is to get an actual cost estimate from the marketing team for outreach versus a missed loan, then re-tune the cutoff to minimize expected dollar loss instead of picking a threshold by eye. The second is to try a tree-based model like random forest or XGBoost, since logit and lasso both assume linearity in the log-odds and acceptance probably has thresholds and interactions those models cannot pick up. We would also want to pull data from earlier campaigns to check whether the patterns hold up over time. Finally, if the bank has behavioral data like online banking activity or app logins, bringing that in would probably help with customers who do not have obvious income or CD signals.

