---
title: "Team Money Dataset"
subtitle: "Pacific Federal Loan Campaign"
author: "Charlie, Arya, Steven, Ryan (Team Money)"
format:
  html:
    df-print: paged
    number-sections: true
    toc: true
    toc-depth: 2
    code-tools: true
    code-overflow: wrap
    embed-resources: true
execute:
  echo: true
  warning: false
  message: false
  cache: false
  error: false
---

```{r}
# Load libs
library(tidyverse)

# set seed
set.seed(310)

# Printing / numeric display
options(width = 70)
options(scipen = 99)

# Chunk defaults (optional, but keeps figures consistent)
knitr::opts_chunk$set(
  fig.width  = 5,
  fig.height = 4,
  fig.align  = "center"
  )
```

# Business Problem

Pacific Federal is a regional bank whose liability (deposit) customers outnumber its asset (loan) customers. The retail banking division recently ran a campaign offering personal loans to its deposit holders, but fewer than 10% accepted. The Chief Retail Officer wants to understand what distinguishes customers who accepted the loan from those who didn't, and build a system to identify high-potential targets for the next campaign round.

# Load the Data

```{r}
loans <- read_csv("data/pacific_federal_loan_campaign.csv")
```

# Overview

```{r}
# Dimensions: rows x columns
dim(loans)
```

```{r}
glimpse(loans)
```

```{r}
head(loans)
```

# Summary Statistics

```{r}
summary(loans)
```

-   9.6% of users accepted the loan offer

-   Average household size is around 2.396

    -   Vast majority are couples with 0-1 kids

-   10.4% of users have a securities account

-   6% of users have a CD

-   29.4% of users have an active credit card

# Data Cleaning

```{r}
loans_clean <- 
  loans %>% 
  mutate(
    Education = factor(Education, 
                       levels = c(1, 2, 3),
                       labels = c("Undergraduate", "Graduate", "Professional")),
    Personal_Loan = factor(Personal_Loan),
    Securities_Account = factor(Securities_Account),
    CD_Account = factor(CD_Account),
    Online = factor(Online),
    CreditCard = factor(CreditCard),
    Income_Per_Family_Member = (Income / Family), 
    Monthly_Income = Income * 12,
    Monthly_CC_Spending_to_Income = CCAvg / Monthly_Income 
  )
```

-   Education changed to factor with labels in order to make visualizations with this predictor easier to interpret

-   Income_Per_Family_Member created to find contribution of each member (gives insight into how much is going to expenses if family is larger)

-   Monthly_CC_Spending_to_Income created to see how much of monthly income goes towards credit card

# Model

```{r}
library(caret)
library(rsample)

set.seed(310)
loans_split <- initial_split(loans_clean, prop=0.8)
loans_train <- training(loans_split)
loans_test  <- testing(loans_split)
logit1 <- glm(Personal_Loan ~ Experience + Income + Education + Monthly_CC_Spending_to_Income + CD_Account + CreditCard + Income_Per_Family_Member,
              data=loans_train, 
              family = binomial)

summary(logit1)

```

-   Mortgage removed due to low significance

-   Age removed due to likely high correlation with experience

# Predict

```{r}

scores_train <- predict(logit1, newdata=loans_train, type="response")
results_train <- 
  loans_train %>% 
    mutate(scores_tr = scores_train)

scores_test <- predict(logit1, newdata=loans_test, type="response")
results_test <- 
  loans_test %>% 
    mutate(scores_te = scores_test)

```

# Report Results

```{r}
pred_class <- ifelse(scores_test > 0.63, 1, 0)


actual <- factor(loans_test$Personal_Loan,
          levels = c(0,1))

predicted <- factor(pred_class,
              levels = c(0,1))

cm <- confusionMatrix(predicted,
      actual,
      positive = "1")

cm
```

-   Initially started with a 90% cutoff because acceptance rate is around 10%

    -   Lowered gradually until 0.62

# Visualizations

```{r}
library(ggplot2)

# Visualization 1: This visualization examines the relationship between annual income and the propensity to accept a personal loan, segmented by education level.

ggplot(loans_train, aes(x = Income, fill = as.factor(Personal_Loan))) +
  geom_density(alpha = 0.5) +
  facet_wrap(~Education) +
  labs(title = "Income Distribution by Loan Acceptance",
       subtitle = "Faceted by Education Level",
       x = "Annual Income (Thousands $)",
       fill = "Personal Loan") +
  theme_minimal()


# Visualization 2: This plot visualizes the predicted probabilities against income, which we believe could be a very significant predictor. 
ggplot(results_train, aes(x = Income, y = scores, color = Personal_Loan)) +
  geom_point(alpha = 0.6) +
  geom_hline(yintercept = 0.5, color="red", linetype="dashed") + 
  labs(title = "Predicted Loan Probability vs. Income",
       x = "Income",
       y = "Predicted Probability of Accepting Loan") +
  theme_minimal()
```

# Plan

1.  More in depth data cleaning (looking at data for outliers and removing them along with missing values)
2.  Adjusting model to find predictors that have the largest prediction impact (rather than just using variables we think may be useful)
3.  Experiment with numerous types of models to find which one best pertains to our question and dataset
4.  Predicting the probabilities of a certain person accepting the loan offer or not
5.  Deciding what metrics are most useful to measure how impactful our model is
6.  Creating visualizations that paint a picture to stakeholders of the type of clients they should be targeting with the new campaign
7.  Refining our model(s)

# Roadblocks/Questions

Deciding the type of model we want to use has been a roadblock. We will continue using what we learned in class (such as random forests) to experiment. Our preliminary model is a logistic regression.
