---
title: "Team Money Dataset"
subtitle: "Pacific Federal Loan Campaign"
author: "Charlie, Arya, Steven, Ryan (Team Money)"
format:
  html:
    df-print: paged
    number-sections: true
    toc: true
    toc-depth: 2
    code-tools: true
    code-overflow: wrap
    embed-resources: true
execute:
  echo: true
  warning: false
  message: false
  cache: false
  error: false
---

```{r}
# Load libs
library(tidyverse)

# set seed
set.seed(310)

# Printing / numeric display
options(width = 70)
options(scipen = 99)

# Chunk defaults (optional, but keeps figures consistent)
knitr::opts_chunk$set(
  fig.width  = 5,
  fig.height = 4,
  fig.align  = "center"
  )
```

# Business Problem

Pacific Federal is a regional bank whose liability (deposit) customers outnumber its asset (loan) customers. The retail banking division recently ran a campaign offering personal loans to its deposit holders, but fewer than 10% accepted. The Chief Retail Officer wants to understand what distinguishes customers who accepted the loan from those who didn't, and build a system to identify high-potential targets for the next campaign round.

# Load the Data

```{r}
loans <- read_csv("data/pacific_federal_loan_campaign.csv")
```

# Overview

```{r}
# Dimensions: rows x columns
dim(loans)
```

```{r}
glimpse(loans)
```

```{r}
head(loans)
```

# Summary Statistics

```{r}
summary(loans)
```

-   9.6% of users accepted the loan offer

-   Average household size is around 2.396

-   10.4% of users have a securities account

-   6% of users have a CD

-   29.4% of users have an active credit card

# Data Cleaning

```{r}
loans_clean <- 
  loans %>% 
  filter(Experience >= 0) %>%
  mutate(
    Education = factor(Education, 
                       levels = c(1, 2, 3),
                       labels = c("Undergraduate", "Graduate", "Professional")),
    Personal_Loan = factor(Personal_Loan),
    Securities_Account = factor(Securities_Account),
    CD_Account = factor(CD_Account),
    Online = factor(Online),
    CreditCard = factor(CreditCard),
    Income_Per_Family_Member = (Income / Family), 
    Monthly_Income = Income * 12,
    Monthly_CC_Spending_to_Income = CCAvg / Monthly_Income,
    100k_Income = ifelse(Income >= 100, 1, 0)
  )
```

-   Entries with negative years of experience removed

-   Education changed to factor with labels in order to make visualizations with this predictor easier to interpret

-   Income_Per_Family_Member created to find contribution of each member (gives insight into how much is going to expenses if family is larger)

-   Monthly_CC_Spending_to_Income created to see how much of monthly income goes towards credit card

# Data Splitting

```{r}
library(caret)
library(rsample)

set.seed(310)
loans_split <- initial_split(loans_clean, prop=0.8)
loans_train <- training(loans_split)
loans_test  <- testing(loans_split)
```

# Model 1

```{r}
library(caret)
library(rsample)

set.seed(310)
loans_split <- initial_split(loans_clean, prop=0.8)
loans_train <- training(loans_split)
loans_test  <- testing(loans_split)
logit1 <- glm(Personal_Loan ~ Experience + Income + Education + Family + CD_Account,
              data=loans_train, 
              family = binomial)

summary(logit1)

```

-   Mortgage removed due to low significance

-   Age removed due to likely high correlation with experience

# Predict Model 1

```{r}

scores_train <- predict(logit1, newdata=loans_train, type="response")
results_train <- 
  loans_train %>% 
    mutate(scores_tr = scores_train)

scores_test <- predict(logit1, newdata=loans_test, type="response")
results_test <- 
  loans_test %>% 
    mutate(scores_te = scores_test)

```

# Report Results Model 1

```{r}
pred_class <- ifelse(scores_test > 0.63, 1, 0)


actual <- factor(loans_test$Personal_Loan,
          levels = c(0,1))

predicted <- factor(pred_class,
              levels = c(0,1))

cm <- confusionMatrix(predicted,
      actual,
      positive = "1")

cm
```

-   Initially started with a 90% cutoff because acceptance rate is around 10%

    -   Lowered gradually until 0.62

# ROC/AUC Model 1 

```{r}
library(pROC)
roc <- roc(actual, scores_test)

# AUC

auc(roc)

# Plot ROC Curve

plot(roc)
```

# Visualizations

```{r}
library(ggplot2)

# Visualization 1: Probability of accepting a personal loan by education level

ggplot(loans_train,
       aes(x = Education,
           fill = as.factor(Personal_Loan))) +
  geom_bar(position = "fill") +
  scale_y_continuous(labels = scales::percent) +
  labs(
    title = "Probability of Accepting a Personal Loan by Education Level",
    x = "Education Level",
    y = "Probability",
    fill = "Personal Loan"
  ) +
  theme_minimal()


# Visualization 2: This plot visualizes the predicted probabilities against income, which we believe could be a very significant predictor. 
ggplot(results_train, aes(x = Income, y = scores_train, color = Personal_Loan)) +
  geom_point(alpha = 0.6) +
  geom_hline(yintercept = 0.5, color="red", linetype="dashed") + 
  labs(title = "Predicted Loan Probability vs. Income",
       x = "Income",
       y = "Predicted Probability of Accepting Loan") +
  theme_minimal()

# Visualization 3: This plots visualizes the amount of users that took the personal loan based on if they had a securities account, cd, or credit card.

library(dplyr)

plot_data <- data.frame(
  Type = c("Securities Account", "CD Account", "Credit Card", "All Three"),
  Percent = c(
    sum(loans_clean$Personal_Loan == 1 & loans_clean$Securities_Account == 1) /
      sum(loans_clean$Securities_Account == 1),
    sum(loans_clean$Personal_Loan == 1 & loans_clean$CD_Account == 1) /
      sum(loans_clean$CD_Account == 1),
    sum(loans_clean$Personal_Loan == 1 & loans_clean$CreditCard == 1) /
      sum(loans_clean$CreditCard == 1),
    sum(loans_clean$Personal_Loan == 1 &
          loans_clean$Securities_Account == 1 &
          loans_clean$CD_Account == 1 &
          loans_clean$CreditCard == 1) /
      sum(loans_clean$Securities_Account == 1 &
            loans_clean$CD_Account == 1 &
            loans_clean$CreditCard == 1)
  )
)
ggplot(plot_data, aes(Type, Percent, fill = Type)) +
  geom_col() +
  geom_text(aes(label = scales::percent(Percent)), vjust = -0.3) +
  scale_y_continuous(labels = scales::percent) +
  labs(
    title = "Percent of Users Taking Personal Loans",
    x = "",
    y = "Percent Taking Loan"
  ) +
  theme_minimal()

```

# Model 2

```{r}
library(glmnet)
library(caret)
library(dplyr)

x_train <- model.matrix(
  Personal_Loan ~ Experience + Income + Education +
  Monthly_CC_Spending_to_Income + CD_Account + Income_Per_Family_Member
  + CreditCard,
  data = loans_train )[, -1]

y_train <- loans_train$Personal_Loan

x_test <- model.matrix(
  Personal_Loan ~ Experience + Income + Education +
  Monthly_CC_Spending_to_Income + CD_Account + Income_Per_Family_Member
  + CreditCard + 100k_Income,
  data = loans_test
)[, -1]

lasso_model <- cv.glmnet(
  x_train,
  y_train,
  family = "binomial",
  alpha = 1
)

coef(lasso_model, s = "lambda.min")


```

# Predict Model 2

```{r}
lasso_probs <- predict(
  lasso_model,
  s = "lambda.min",
  newx = x_test,
  type = "response"
)
```

# Report Results Model 2

```{r}

lasso_pred <- ifelse(lasso_probs > 0.6, 1, 0)

actual_lasso <- factor(loans_test$Personal_Loan,
                 levels = c(0,1))

predicted_lasso <- factor(lasso_pred,
                    levels = c(0,1))

cm_lasso <- confusionMatrix(predicted_lasso,
                      actual_lasso,
                      positive = "1")

cm_lasso

```

# ROC/AUC Model 2

```{r}

library(pROC)

roc_lasso <- roc(actual_lasso, as.numeric(lasso_probs))
auc(roc_lasso)
plot(roc_lasso)

```



# I have also added some things to the powerpoint 
# Re model evaluation, we don't really gaf much about false positives (lower outreach costs) compared to false negatives (foregone interest on loans)
# How are we doing an AI log for this


